JD notes on effects

### 2022-07-04 manuscript/body.tex
No major progress, we might need to discuss first.

Plots of predicted values of an outcome against predictors (often called effects plots, or prediction plots) are often a useful way to summarize the results of a regression model. These can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change. In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}. \jd{Do these citations make sense where I put them?}

@@

\jd{How much do we need of this \P? To what extent can it be merged into the two above.}

\BC{Added a sentence in the last \P. I think we can delete the \P}

\jd{We need to stake out what we need by predictions and effects first. We could do this by first talking about the CI choices, and about univariate vs. multivariate? Or by talking about what people might want from predictions and effects first? We can't imply that this stuff is already known.}

@@

\dya{Both predictions and effects tell us what we expect the response to be for a given value of \emph{focal} input variable (scientific variable of interest). The major difference between the two lies in the construction of the associated confidence intervals. In particular, for predictions, we want to use the conventional curved confidence intervals (or prediction intervals) which capture the uncertainty in the whole model and generally focus on the \emph{total} effects, and so will often be suited for a univariate model. On the other hand, effects, attempt to show the effect of a focal input variable. In this case, we suggest the narrower confidence intervals which focus on uncertainty in the effect of the focal input variable only, and so will often be suited for the direct effects – after controlling for other non-focal input variables – from a multivariate model.}

\dya{When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret \jd{Keep working on reducing repetition if you can.}.}

@@

\BC{I think we should move all with "MOVE" tag to somewhere after the statistical background but Ben suggested this format. I think moving everthing downstream will streamline the flow.}
\MOVE{The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively. However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}). For simple OLS models (with no interactions), the two approaches give the same predictions. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
\end{align}
%
}

\MOVE{We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the predictions based on the three approaches match the true predictions. B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}
%
In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$. In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.
}

\MOVE{One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects. However, currently, the two packages do not provide a straightforward way do achieve this. To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above. The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}. For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point. For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice). Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor. The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor. Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.
}

@@

\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors.combined.pdf}
\end{center}
\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero? Also, the real difference between anchors is for multi-parameter predictors? Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor. As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines). \MOVE{Should me move this downstream}} 
\label{fig:justify_anchors}
\end{figure}
%
\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}. The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects. To illustrate the distinction between prediction and effects plots, and marginal effects, we use the simulated data described in \nameref{S1_Appendix}. 
%
\begin{figure}
\centering
\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
\caption{{\bf A comparison of effects and marginal effect plots.} \NEW{A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval. B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals. For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects. However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.} \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions. Maybe you should consider making a box? Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.} \MOVE{I think this should also be moved downstream.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
\label{fig:qoi_age_pred_plot}
\end{figure}
%

@@

\jd{Are we kind of going backward now from the comparison section? Should all of these definitions come first?}
\BC{I agree. We can move all the package comparisons to simulation section. \MOVE{}}

\jd{This \P\ seems unnecessary. Why not just say “To illustrate \ldots", and then jump to the next \P (“, we consider")}

\NEW{To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$. Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$. Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the \fix{central (outcome) estimate} with respect to the focal input variable in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.
}

@@

>>>>

Plots of predicted values of an outcome against predictors (often called effects plots, or prediction plots) are often a useful way to summarize the results of a regression model. These can be used to illustrate model uncertainty, or to give a more explicit quantitative sense of how the outcome is expected to change. In generalized models with a non-linear link function, or models with a spline or polynomial response to a predictor variable, they can also aid in undertanding difficult-to-interpret coefficient estimates \citep{brambor_understanding_2006, berry_improving_2012, leeper2017interpreting}.

@@

@@

@@

\BC{I think we should move all with "MOVE" tag to somewhere after the statistical background but Ben suggested this format. I think moving everthing downstream will streamline the flow.}

@@

\item \textbf{Marginal effects:} In simple linear models with no interaction terms, the coefficient estimates are simple and directly interpretable as the expected change in outcome for a unit change in focal input variable. This is the \emph{unconditional marginal effect} and it is constant across all the observations and levels of all other input variables. Consider models described by Equation~\ref{eq:simple_inter_higher_no_interaction} and Equation~\ref{eq:simple_inter_higher}. The marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher_no_interaction} is $\frac{\partial y}{\partial x_2} = \beta_2$. On the other hand, the marginal effect of $x_2$ in model~\ref{eq:simple_inter_higher} is given by $\frac{\partial y}{\partial x_2} = \beta_2 + \beta_{23}x_3$. In other words, if there are no interactions, the marginal effect of $x_2$ on $y$ is constant, while, if there are interactions in the model, the marginal effect of a change in $x_2$ on $y$ depends on the value of the other \emph{conditioning} predictor, $x_3$. In essence, computation of marginal effects requires the use of partial derivatives and can be considered as the slope multi-dimensional surface with response to the surface of interest \cite{leeper2017interpreting}. 

@@

To illustrate mean-based approach, consider a simple linear model with linear predictor $\eta = \bX\bbeta$ and let $g(\boldmu) = \boldeta$ be an identity link function, where $\boldmuh$ is the expected value of response variable $\hat{y}$. Let $\bbetah$ be the estimate of $\bbeta$, together with the estimated covariance matrix $\Sigma = V(\bbetah)$ of $\bbetah$. Then the quantity $\hat{\boldeta}^\star = \bX^\star\bbetah$ is the \fix{central (outcome) estimate} with respect to the focal input variable in question \citep{fox2009effect}; where $\bX^\star$ is the centered model matrix constructed by averaging the columns of non-focal predictors in model matrix $\bX$, and together with appropriately chosen values of focal predictor.

@@

\subsection{Comparison with other implementations}

The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively. However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}). For simple OLS models (with no interactions), the two approaches give the same predictions. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:
%
\begin{align}
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \epsilon \label{eq:simple_inter_higher_no_interaction}\\
y &= \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_{23}x_2x_3 + \epsilon \label{eq:simple_inter_higher}.
\end{align}
%

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.inter.pdf}
\end{center}
\caption{{\bf A comparison of \pkg{emmeans}, \pkg{effects} and \pkg{varpred} predictions for regression of $x_1$ on $y$.} The horizontal dashed lines are the mean of the predicted and simulated $y$, $\bar{\hat{y}}$ and $\bar{y}$, respectively. The grey points are the binned simulated $y$. The trend lines represent the corresponding $\hat{y}$ at various levels of $x_1$, while holding the other input variables at their mean and the true predictions based on the simulation parameters. A: In the absence of interaction, the predicted mean, $\bar{\hat{y}}$, closely matches the simulated value in all the three approaches, i.e., $\bar{y} \approx \bar{\hat{y}}$. Similarly, the predictions based on the three approaches match the true predictions. B: Even with the simple interaction between non-focal input variables, we start seeing deviation in both predicted mean, $\bar{\hat{y}}$, and overall predictions from the simulated, $\bar{y}$, and true predictions, respectively, in two commonly used packages (\pkg{emmeans} and \pkg{effects}), but not the proposed \pkg{varpred}.}
\label{fig:justify_plots}
\end{figure}
%
In the absence of interactions (\EREF{simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interactions, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates start to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular, \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$. In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.

One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects. However, currently, the two packages do not provide a straightforward way do achieve this. To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}. 
%
\begin{figure}
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf The prediction and effect styled plots at $95\%$ confidence intervals.} The description of horizontal, vertical and trend lines remain the same as above. The wider dashed curves corresponds to the conventional confidence intervals around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower curves crossing at the mean of the focal input variable, i.e., the center point, correspond to the effect of the focal input variable from \pkg{varpred}. For simple OLS models, the effect-styled confidence bands crosses at the center point (an anchor). The prediction-styled curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables. On the other hand, effect-styled curves only take into account the main effect uncertainty of the focal input variable.}
\label{fig:justify_ci_plots}
\end{figure}
%
For \pkg{emmeans} and \pkg{effects}, the confidence intervals are much wider because they include uncertainties associated with the intercept and non-focal input variables, but narrower and crosses at the mean of the focal input variable in \pkg{varpred}. In other words, with \pkg{varpred}, we are able to generate effects indicating zero uncertainty at the value of the focal input variable we are more certain about, i.e., the center point. For simple OLS models, the point where the confidence intervals curves cross is the model center or simply center point; it corresponds to anchor we choose, in this case, the center point (which is generally an appropriate choice). Predictions are the standard curves; they tell us what we expect to see if we know the value of the focal predictor. The lines inside (effects) are a little different; they are a good way to visualize the effect of the focal predictor. Unlike predictions, where the uncertainty captures the uncertainty in the whole model, the effects plot focuses on uncertainty in the effect of the focal predictor only and depend on the anchor.

\jd{Why do we need this in the definitions section?:} In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the center (mean) of the focal input variable with \ldots.
%
\begin{figure}
\begin{center}
\includegraphics{justify_anchors.combined.pdf}
\end{center}
\caption{{\bf A comparison of zero- and center-anchored effects.} \jd{This explanation is not currently correct, but do we really need to explain that the zero anchor means that the anchor is at zero? Also, the real difference between anchors is for multi-parameter predictors? Do we have an example of that?} The center-anchored effects are based on the mean of the focal predictor while the zero-anchored effects are based on the minimum of the focal predictor. As we mentioned earlier, the choice of the anchor do not affect the main estimates (central trend lines).} 
\label{fig:justify_anchors}
\end{figure}
%

The \proglang{R} software package \pkg{margins} \cite{lenth2018package} generates marginal effects. To illustrate the distinction between prediction and effects plots, and marginal effects, we use the simulated data described in \nameref{S1_Appendix}. 
%
\begin{figure}
\centering
\includegraphics{cubic_varpred_margins_pred.ggp.pdf}
\caption{{\bf A comparison of effects and marginal effect plots.} \NEW{A:  The black solid line is the central (outcome) estimate, while the dashed lines corresponds to the effects styled $95\%$ confidence interval. B: The marginal effect (ME) at representative values of the focal input variable \code{age} together with the $95\%$ confidence intervals. For simple linear models with no interactions, the ME should be constant across all the values of the focal input variable; and it would correspond to the slope of the central estimate of the prediction or effects. However, in this example, the focal input variable is a cubic polynomial, and so, the effect is not constant.} \jd{The definitions should be shorter and it's definitely weird to have a figure in the definitions. Maybe you should consider making a box? Also, if we're going to have this figure, we should try to illustrate a more difficult example for comparison.} \BC{What should we call the central and/or y-axis line of the effect/prediction plot?}}
\label{fig:qoi_age_pred_plot}
\end{figure}
%
